---
title: "DATA_605_Assignment_1_Thonn"
file: GIyengar_DATA_605_Assign1_Thonn
name: Daniel Thonn
output:
  html_document: default
  pdf_document: default
course: DATA 605 CUNY MSDA program
---

** DATA_605_Assignment_1_Thonn **

** Problem Set 1 **

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

(1) Calculate the dot product u.v where u = [0.5; 0.5] and v = [3;-4]
``````{r, echo=TRUE}
u <- c(0.5, 0.5)

v <- c(3, -4)

dotprod1 <- u %*% v

dotprod1

```

(2) What are the lengths of u and v?
``````{r, echo=TRUE}

lengthu <- sqrt(u %*% u)
lengthu

lengthv <- sqrt(v %*% v)
lengthv

```

(3) What is the linear combination: 3u - 2v ?
``````{r, echo=TRUE}

lincombination1 <- 3*u - 2*v

lincombination1


```

(4) What is the angle between u and v ?  (radians)
``````{r, echo=TRUE}
angle_rad <- acos((u %*% v)/(sqrt(u %*% u)*sqrt(v %*% v)))
angle_rad

```

** Problem Set 2 **

Set up a system of equations with 3 variables and 3 constraints and solve for x. Please write a function in R that will take two variables (matrix A & constraint vector b) and solve using elimination. Your function should produce the right answer for the system of equations for any 3-variable, 3-equation system. You don't have to worry about degenerate cases and can safely assume that the function will only be tested with a system of equations that has a solution. Please note that you do have to worry about zero pivots, though. Please note that you should not use the built-in function solve to solve this system or use matrix inverses. The approach that you should employ is to construct an Upper Triangular Matrix and then back-substitute to get the solution. Alternatively, you can augment the matrix A with vector b and jointly apply the Gauss Jordan elimination procedure.

Please test it with the system below and is should produce a solution
x = [-1.55,-0.32,0.95]

A = matrix(c(1, 2, -1, 1, -1, -2, 3, 5, 4), nrow=3, ncol=3)
b = c(1, 2, 6)


``````{r, echo=TRUE}

# Function 1 - swap elements - vector
elements_swap <- function(v, m, n){
  a_v <- v[m]
  v[m] <- v[n]
  v[n] <- a_v
  return(v)
}

# Function 2 - swap rows - matrix
rows_swap <-function(A, m, n){
  a  <- A[m,]
  A[m,] <- A[n,]
  A[n,] <- a
  return(A)
}

# Function 3 - find pivot
pivot_find <- function (A, r, c){
# Initialize variables
  swap_a <- FALSE
  pv <- 0
  swap_r <- 0
  loop1 <- TRUE
  j <- r
  #loop thru each matrix and locate non-zero pivot point for column c
  while(loop1){
    if(A[j,c] != 0){
      loop1 <- FALSE
      pv<-A[j,c]
      if(j>r){
        swap_r <- j
        swap_a <- TRUE
      }
    }
    j <- j + 1
    if (j > nrow(A)){
      loop <- FALSE
    }
  }
  return(list(pv, swap_a, swap_r))
} 

# Main function to Solve equation (3x3)
solve_main <- function (A, b){
  # Check input parameters: A, b, and whether row(A) equals length(b)
  if(missing(A)){
    stop("matrix A not in form Ax = b")
  }
  if(missing(b)){
    stop("vector b not in form Ax = b")
  }
  if(nrow(A) != ncol(A)){
    stop("A not a square matrix")
  }
  if(nrow(A) != length(b)){
    stop("number of rows of A != dim of b")
  }
  num_rows <- nrow(A)
  num_col <- ncol(A)
  A1 <- A
  b1 <- b
  # Create Upper Triangular Matrix 
  # Initialize column cursor
  c <- 1
  while (c <= num_col){
    # Initialize row cursor
    r <- c
    # find pivot point
    l_result <- pivot_find (A1, r, c)
    p <- unlist(l_result[1])
    swap_indicator <- unlist(l_result[2])
    swap_row <- unlist(l_result[3])
    # Validate that pivot point exists
    if (p==0){
      text <- "No pivot point found in column"
      txt_col <- as.character(c)
      msg <- paste(text,txt_col)
      stop(msg)
    }
    # if pivot point check for swap
    if(swap_indicator){
      A1 <- rows_swap(A1,r,swap_row)
      b1 <- elements_swap (b1, r, swap_row)
    }
    # Process elimination for each row
    k <- r + 1
    while (k <= num_rows){
      # multiplier check
      m = A1[k,c]/p
      A1[k, ] <- A1[k, ] - m*A1[r, ]
      b1[k] <- b1[k] - m*b1[r]
      k <- k + 1
    } # end loop rows
    c <- c + 1
  } # end loop columns
  
  # Solve equations
  equation_v <- vector()
  i <- num_col
  equation_v[i] <- b1[i]/A1[i,i]
  z<- b1[3]/A1[3,3]
  y<- (b1[2] - A1[2,3]*z)/A1[2,2]
  x<- (b1[1] - A1[1,2]*y - A1[1,3]*z)/A1[1,1]
  result1 <- c(round(x, digits = 2), round(y, digits = 2), round(z, digits = 2))
  return(result1)
}

# Test the Main Function #
A = matrix(c(1, 2, -1, 1, -1, -2, 3, 5, 4), nrow=3, ncol=3)
b = c(1, 2, 6)
result <- solve_main(A,b)
# actual result:
result
# expected result: [1] -1.55 -0.32  0.95

```


**END**
